---
title: "Symptom-Based Disease Prediction"
author: "Trevor Okinda"
date: "2025"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | A Symptom-Based disease prediction model |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/miltonmacgyver/symptom-based-disease-prediction-dataset\>*

### Reference:

*\*MiltonMacGyver. (n.d.). Symptom-Based Disease Prediction Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/miltonmacgyver/symptom-based-disease-prediction-dataset*
\>\

Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load libraries
library(tidyverse)

# Choose and load dataset with correct column types
file_path <- file.choose()

SymptomData <- read.csv(file_path, colClasses = c(
  fever = "factor",
  cough = "factor",
  headache = "factor",
  nausea = "factor",
  vomiting = "factor",
  fatigue = "factor",
  sore_throat = "factor",
  chills = "factor",
  body_pain = "factor",
  loss_of_appetite = "factor",
  abdominal_pain = "factor",
  diarrhea = "factor",
  sweating = "factor",
  rapid_breathing = "factor",
  dizziness = "factor",
  label = "factor" # Target variable
))

# Quick check
str(SymptomData)
summary(SymptomData)

# Frequency of each disease label
table(SymptomData$label)
```
## Measures of Frequency
```{r MOF}
# Frequency of each disease label
table(SymptomData$label)

# Frequency of each symptom being present (1) or absent (0)
symptom_freq <- sapply(SymptomData[ , -ncol(SymptomData)], function(col) table(col))
symptom_freq
```

## Measures of Distribution
## Measures of Relationship
## Plots





